---
title: "F and M formula effects: recovery and permitted combinations"
output:
  html_document:
    self_contained: true
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 120, fig.path = "results/figures/")
results <- readRDS("results/mean_effects_recovery.rds")
simple <- do.call(rbind, results$simple)
full <- do.call(rbind, results$full)
```

## Main findings

**Retain the restricted mean-formula options, but do not treat them as generally
validated for every assessment.** All 600 isolated fits passed the numerical
checks. The 360 full assessment fits were more demanding: numerical success
ranged from 67% to 97% across specifications. No tolerances were relaxed and
no fit was rescued.

- **F:** mean-process SDs were reasonably recovered. All 90 F fits had optimizer
  code zero and a positive-definite Hessian; 17 missed the gradient threshold.
  RW had the most such failures (10/30). This deserves attention in applications,
  but is not evidence by itself that the process density is wrong.
- **M:** a small shared signal and IID residuals were difficult to separate.
  AR1/RW residual SDs tended to be underestimated, with wide intervals and
  occasional non-positive-definite Hessians. Increasing the shared SD from 0.12
  to 0.30 brought median residual SDs close to the generating 0.06, although
  some Hessian failures remained. Start with a structured M mean alone; add
  residual variation only when the data or external information support it.
- **AR1 persistence:** correlation tended to be underestimated and intervals
  were wide, including in the isolated experiment. Do not interpret a fitted
  correlation as a precise description of persistence without a sensitivity
  check.

The deterministic tests independently match the marginal Gaussian likelihood
to an exact calculation. That establishes the density mathematics; the recovery
experiments reveal the remaining numerical and information limits.

## What is being tested?

Can a shared change in mortality be separated from independent age-specific
fluctuations? Formula terms describe variation in the **mean log F or log M**.
The absolute latent mortality states remain unchanged in meaning.

The study uses `iid(year)`, `rw(year)` and `ar1(year)`, with estimated process
SDs and estimated AR1 correlation. IID effects fluctuate independently; RW
changes persist; AR1 fluctuations return towards the baseline. RW starts at
zero. These assumptions have proper, distinct process densities.

The simpler experiment has 40 years with eight replicated ages, or only two
ages. It observes log mortality with known measurement SD 0.05 and separately
estimated IID residual SD 0.10. True mean-process SD is 0.18 and AR1 correlation
is 0.65. There are `r results$n_simple` replicates per design and process.

The full experiment has 30 years, eight ages, catch at age, and two surveys at
different times of year. It estimates recruitment, age-specific mean F, F
residual variability, and survey q. M's baseline is supplied. Observation SDs
are known (0.08), recruitment SD is 0.12, F residual SD is 0.08, M residual SD is
0.06 when enabled, and shared mean-process SD is 0.12. There are
`r results$n_full` replicates per specification. M is tested with and without
IID residuals. This deliberately informative study is **not a validation of
every real assessment or sparse-data situation**.

A second M experiment increases the shared mean-process SD to 0.30 while
retaining IID residual SD 0.06. This tests whether poor separation of a small
M signal improves when there is more information about its variation. SD is
marginal for IID, increment SD for RW and innovation SD for AR1.

Full fits start at the generating fixed coefficients and absolute states,
but reset the mean trajectory to zero, process SDs to 0.10, and AR1 correlation
to 0.50. These favorable starts isolate recovery and numerical problems; they
do not establish that arbitrary starts work. No failed fit is rescued or
removed from success-rate denominators.

## Numerical reliability

```{r reliability}
numerical <- function(tab, groups, n) {
  first <- tab[!duplicated(tab[, c(groups, "replicate")]), ]
  do.call(rbind, lapply(split(first, interaction(first[groups], drop = TRUE)), function(d) {
    good <- d$optimizer == 0 & d$pdHess & is.finite(d$gradient) & d$gradient <= .01
    if ("converged" %in% names(d)) good <- d$converged
    cbind(d[1, groups, drop = FALSE], attempted = n, returned = nrow(d),
      success = sum(good), success_percent = round(100 * sum(good) / n, 1),
      optimizer_ok = sum(d$optimizer == 0), gradient_ok = sum(is.finite(d$gradient) & d$gradient <= .01),
      Hessian_ok = sum(d$pdHess), max_gradient = max(d$gradient))
  }))
}
simple_rates <- numerical(simple, c("type", "design"), results$n_simple)
full_rates <- numerical(full, c("component", "type", "design"), results$n_full)
knitr::kable(full_rates[, c("component", "type", "design", "success",
  "optimizer_ok", "gradient_ok", "Hessian_ok", "max_gradient")],
  digits = 3, row.names = FALSE,
  col.names = c("Component", "Process", "Design", "Passed", "Optimizer",
                "Gradient", "Hessian", "Max gradient"))
```

A successful fit requires optimizer code zero, maximum absolute gradient at
most 0.01 and a positive-definite fixed-effect Hessian. These checks establish
numerical reliability, not biological identifiability. The simpler experiment
uses the same thresholds. Each full specification has 30 attempted fits;
every attempted fit returned a result.

The separate counts distinguish optimizer failures, an insufficiently small
gradient, and uncertainty failures. A missed gradient threshold alone does not
establish structural confounding; poor variance recovery or repeated boundary
estimates give additional evidence.

```{r reliability-figure}
rates <- rbind(data.frame(label = paste("Isolated", simple_rates$type, simple_rates$design),
                          percent = simple_rates$success_percent),
               data.frame(label = paste(full_rates$component, full_rates$type, full_rates$design),
                          percent = full_rates$success_percent))
par(mar = c(4, 16, 1, 1))
barplot(rates$percent, names.arg = rates$label, horiz = TRUE, las = 1,
        xlim = c(0, 100), col = "steelblue", cex.names = .8,
        xlab = "Numerically successful fits (%)")
abline(v = 90, lty = 2)
```

## Are the process components recovered?

```{r recovery-table}
recovery <- function(tab, groups) {
  do.call(rbind, lapply(split(tab, interaction(tab[groups], drop = TRUE)), function(d) {
    available <- is.finite(d$estimate)
    intervals <- is.finite(d$lower) & is.finite(d$upper) & d$pdHess
    cbind(d[1, groups, drop = FALSE], n_estimates = sum(available),
      true_value = d$truth[1], median_estimate = median(d$estimate[available]),
      mean_bias = mean(d$estimate[available] - d$truth[available]),
      n_intervals = sum(intervals), median_CI_width = median(d$upper[intervals] - d$lower[intervals]),
      coverage = mean(d$lower[intervals] <= d$truth[intervals] &
        d$upper[intervals] >= d$truth[intervals]))
  }))
}
summary_full <- recovery(full, c("component", "type", "design", "parameter"))
knitr::kable(summary_full[, setdiff(names(summary_full), "n_estimates")],
  digits = 3, row.names = FALSE,
  col.names = c("Component", "Process", "Design", "Parameter", "True",
                "Median", "Bias", "Intervals", "CI width", "Coverage"))
```

Coverage is conditional on available intervals with a positive-definite
Hessian; numerical failures must be considered alongside it. With 30
replicates, approximate Monte Carlo uncertainty around nominal 95% coverage is
about ±8 percentage points; with 100 it is about ±4. A single unusual replicate
is not a reason to add constraints or optimization tricks.

```{r recovery-figure, fig.height=8}
par(mfrow = c(1, 2), mar = c(4, 9, 2, 1))
for (component in c("F", "M")) {
  d <- full[full$component == component & full$parameter != "phi", ]
  label <- paste(toupper(d$type), ifelse(d$design == "mean only", "alone",
    ifelse(grepl("strong", d$design), "strong + IID", "+ IID")),
                 ifelse(d$parameter == "mean_sd", "mean SD", "residual SD"))
  label <- factor(label, levels = unique(label))
  boxplot(d$estimate / d$truth ~ label, las = 1, ylim = c(0, 3), horizontal = TRUE,
          xlab = "Estimated SD / true SD", ylab = "", main = component,
          col = "lightblue", outline = FALSE, cex.axis = .8)
  abline(v = 1, lty = 2)
}
```

The figure suppresses extreme outlier points to keep ordinary estimates
readable; the recovery table uses all finite estimates.

```{r correlation, fig.height=3.5}
d <- full[full$parameter == "phi", ]
label <- paste(d$component, ifelse(d$design == "mean only", "mean only",
  ifelse(grepl("strong", d$design), "strong mean + IID", "mean + IID")))
par(mar = c(4, 13, 1, 1))
boxplot(d$estimate ~ label, horizontal = TRUE, las = 1, ylim = c(0, 1),
        xlab = "Estimated AR1 correlation", ylab = "", col = "lightblue")
abline(v = .65, lty = 2)
```

The dashed line is the generating correlation (0.65). In full assessments,
median correlation-interval widths range from 0.55 to 0.80. Wide intervals can
cover the truth while still providing little information about persistence.
Mean-SD interval coverage was 83% for the F-AR1 specification; the study does
not demonstrate uniformly calibrated uncertainty.

## Example trajectories

```{r trajectories, fig.height=7}
examples <- results$examples[grepl("^full_.*_TRUE_1$", names(results$examples))]
par(mfrow = c(2, 3), mar = c(3, 4, 2, 1))
for (d in examples) {
  limits <- range(c(d$truth, d$lower, d$upper), finite = TRUE)
  plot(d$year, d$truth, type = "n", ylim = limits, xlab = "Year", ylab = "Mean log effect",
       main = paste(d$component[1], d$type[1]))
  polygon(c(d$year, rev(d$year)), c(d$lower, rev(d$upper)),
          col = adjustcolor("steelblue", .2), border = NA)
  lines(d$year, d$truth, lwd = 2)
  lines(d$year, d$estimate, col = "steelblue", lwd = 2)
  abline(h = 0, col = "grey")
}
```

Black lines show the generating mean effect; blue lines and bands show the
estimate and its local 95% interval. Examples are the **first replicate**, not
selected for a good fit. The report evaluates mean components, not agreement
with an accepted stock assessment.

Zero means no change from the baseline mortality. A log effect of 0.2 raises
the rate by about 22%; an effect of -0.2 lowers it by about 18%.

```{r stronger-M, fig.height=3.5}
strong <- results$examples[grepl("_strong_1$", names(results$examples))]
if (length(strong)) {
  par(mfrow = c(1, 3), mar = c(3, 4, 2, 1))
  for (d in strong) {
    plot(d$year, d$truth, type = "n", ylim = range(d$truth, d$lower, d$upper, finite = TRUE),
         xlab = "Year", ylab = "Mean log effect", main = paste("Stronger M", d$type[1]))
    polygon(c(d$year, rev(d$year)), c(d$lower, rev(d$upper)),
            col = adjustcolor("steelblue", .2), border = NA)
    lines(d$year, d$truth, lwd = 2)
    lines(d$year, d$estimate, col = "steelblue", lwd = 2)
  }
}
```

## What should be permitted?

| Combination | Treatment |
|:--|:--|
| One shared IID, RW or AR1 mean with IID residuals | Enabled, subject to design checks; interpret recovery results above before estimating both components in a stock assessment. |
| Structured M mean with residual process off | Enabled; supplied M sets the baseline unless an ordinary mean formula estimates it. |
| Numeric `by` | One trajectory multiplied by the covariate; not separate age trajectories. |
| Categorical `by` | Separate trajectories with shared term-level SD/correlation; require adequate replication and compatible M blocks. |
| Estimated IID mean variance duplicating IID residual variance | Rejected; supply one SD or share effects across replicated states. |
| Structured means with RW/AR1 residuals | Not enabled pending a separate identifiability study. |
| Multiple temporal mean terms | Not enabled in this phase. |

Examples of the supported interface are `F_settings = list(process = "iid",
mu_form = ~ factor(age) + rw(year))` and `M_settings = list(process = "off",
mu_form = ~ 0 + ar1(year), mu_supplied = ~ M_assumption)`. The second example
uses supplied M as the baseline and estimates shared deviations around it.

The package has no F residual-process-off option; this work does not add one.
F projections retain the existing terminal-state policy. Mean-process future
states have proper densities and integrate out without observations.

```{r errors}
if (length(results$failures)) {
  knitr::kable(data.frame(run = names(results$failures), reason = unlist(results$failures)),
               caption = "Runs that failed to return a result")
}
```

The study cannot justify unrestricted overlapping temporal components. It also
does not establish robustness when observation errors are unknown, q varies
through time, or N has additional process variation. Those need further
simulations or stock-specific sensitivity analysis. The recovery experiments
vary one mortality mean at a time; jointly time-varying F and M means have not
been evaluated here.

## Reproducibility

Run `simulate_mean_effects.R` from the repository root, then render this report.
Seeds, configurations and all transformations are in that script. The result
object contains the R session information. Generated results, example fits and
this rendered report live in the ignored `results/` folder.

Package validation passed 1,564 assertions, with one interactive-only test
skipped. `R CMD check --no-manual` returned **Status: OK**. The PDF manual check
failed on existing Unicode mathematical symbols in the simulation help page;
its other checks passed.
