---
title: "Process expansion: recovery and assessment trials"
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4.5, fig.path = "results/process_expansion/figures/")
attempts <- readRDS("results/process_expansion/recovery.rds")
trials <- readRDS("results/process_expansion/assessments/trials.rds")
```

## What was tested

F can follow a random walk whose annual changes are correlated across ages.
Ordinary age-based formulas can share or vary process SDs for F, N and M.
Spawning-time SSB accounts for mortality before spawning and supplies the same
parent biomass to stock–recruit models.

The study includes 100 isolated Gaussian replicates for each common/grouped
SD design for F, N and M, and 30 full assessments for each grouped design.
Full assessments use 30 years, eight ages, two surveys and known observation
log-SDs of 0.08. These informative synthetic data do not validate every stock.
All attempts and warnings are retained; no failed fit is rescued.

## Numerical success and recovery

```{r}
groups <- sub("_[0-9]+$", "", names(attempts))
summary <- do.call(rbind, lapply(unique(groups), function(group) {
  a <- attempts[groups == group]
  data.frame(Design = group, Attempts = length(a),
    Passed = sum(vapply(a, function(x) isTRUE(x$numerical), logical(1))),
    Errors = sum(vapply(a, function(x) !is.null(x$error), logical(1))),
    `Hessian passed` = sum(vapply(a, function(x) isTRUE(x$pdHess), logical(1))), check.names = FALSE)
}))
knitr::kable(summary)
recovery <- do.call(rbind, lapply(names(attempts), function(name) {
  d <- attempts[[name]]$recovery
  if (is.null(d)) return(NULL)
  d$design <- sub("_[0-9]+$", "", name)
  d
}))
coverage <- aggregate(covered ~ design + parameter, recovery, function(x) mean(x, na.rm = TRUE))
knitr::kable(coverage, digits = 2, col.names = c("Design", "Parameter", "95% interval coverage"))
```

Coverage uses attempts with available intervals, including numerically failed
fits if intervals were returned. Errors remain in the numerical-success
denominator. SD coefficients are on the log-SD scale. Correlation intervals
use its unconstrained transformation; they are not truncated Normal intervals.

```{r}
full <- recovery[startsWith(recovery$design, "full"), ]
par(mar = c(5, 11, 4, 2) + .1)
for (component in c("F", "N", "M")) {
  d <- full[full$design == paste0("full_", component, "_grouped"), ]
  d$error <- d$estimate - d$truth
  d$parameter <- factor(d$parameter,
    levels = c("(Intercept)", "age_groupyoung", "rho_f"),
    labels = c("Older log SD", "Young/older log SD", "Age correlation"))
  boxplot(error ~ parameter, d, horizontal = TRUE, las = 1,
    main = paste(component, "process: parameter recovery"),
    xlab = "Estimate minus generating value", ylab = "")
  abline(v = 0, col = "grey50", lty = 2)
}
```

All 690 fits passed numerical checks. Full-assessment median older/young SDs
were 0.199/0.101 for F, 0.203/0.097 for N and 0.199/0.085 for M, compared with
generating values of 0.20/0.10. Some young-age M SDs approached zero. Wide
intervals and convergence should not be mistaken for precise M recovery.
F age-correlation interval coverage was 26/30 in full assessments; these
small experiments do not establish generally calibrated uncertainty.

## Source-supported assessment trials

EBS pollock tests the pinned source spawning fraction, 0.25. Northern Shelf
haddock tests correlated F changes and the separate plus-group N SD. North
Sea plaice tests correlated F changes and its four published F-SD sharing
keys. Individual changes precede combinations; combinations are attempted
only when both individual fits pass numerical checks. Existing recipes are
unchanged.

```{r}
trial_summary <- do.call(rbind, lapply(names(trials), function(name) {
  fit <- trials[[name]]$fit
  data.frame(Trial = name, Converged = !is.null(fit) && fit$is_converged,
    `Positive Hessian` = !is.null(fit$sdrep) && isTRUE(fit$sdrep$pdHess),
    Gradient = if (is.null(fit)) NA_real_ else max(abs(fit$gradient)),
    Error = if (is.null(trials[[name]]$error)) "" else trials[[name]]$error, check.names = FALSE)
}))
knitr::kable(trial_summary, digits = 5)
comparisons <- do.call(rbind, lapply(names(trials), function(name) {
  d <- trials[[name]]$comparison_summary
  if (is.null(d)) return(NULL)
  d$trial <- sub("^.*/", "", name)
  d
}))
if (!is.null(comparisons)) {
  selected <- comparisons[comparisons$metric %in% c("ssb", "N", "recruitment", "F"), ]
  knitr::kable(selected[c("assessment_id", "trial", "metric", "comparison_status",
    "mean_absolute_percent_difference", "trend_correlation")], digits = 2)
}
```

Scale disagreement and trajectory agreement are separate. Fixed-M agreement
does not count as improvement. SSB timing can match while biological inputs
remain approximate. Known observation dependence, q powers and selectivity
differences remain outside these additions. Local dashboards and cached
diagnostics provide the detailed review; no assessment specification is
automatically selected from these results.

```{r ssb-trends}
for (id in unique(sub("/.*$", "", names(trials)))) {
  available <- trials[startsWith(names(trials), paste0(id, "/"))]
  reference <- available[[1]]$ref$pop$ssb
  series <- lapply(available, function(x) {
    d <- x$fit$pop$ssb
    scale <- x$ref$comparison_scales[["ssb"]]
    if (is.null(scale)) scale <- 1
    d$est <- d$est * scale
    d
  })
  title <- c(afsc_pollock_ebs_2024 = "Eastern Bering Sea pollock",
    ices_haddock_north_sea_2026 = "Northern Shelf haddock",
    ices_plaice_north_sea_2026 = "North Sea plaice")[[id]]
  plot(reference$year, reference$est, type = "l", lwd = 2,
    ylim = range(c(reference$est, unlist(lapply(series, `[[`, "est"))), na.rm = TRUE),
    xlab = "Year", ylab = "Female SSB (source reporting units)", main = title)
  for (i in seq_along(series)) lines(series[[i]]$year, series[[i]]$est, col = i + 1, lwd = 2)
  labels <- gsub("_", " ", sub("^.*/", "", names(available)))
  legend("topleft", c("Accepted", labels),
    col = seq_len(length(series) + 1), lwd = 2, bty = "n", cex = .8)
}
```

EBS spawning time reduces the SSB scale difference from 39.6% to 27.3%, with
essentially unchanged population states and residuals. Its refit failed the
gradient criterion (0.037 versus 0.01), despite optimizer success and a positive
Hessian. At identical parameters the likelihood is bit-for-bit unchanged:
this recipe has no stock–recruit curve, so spawning time changes reporting
only. The failed restart is retained, not rescued or promoted.

Haddock's grouped N SD improves SSB, N and F agreement but worsens recruitment
agreement; its largest survey residual also increases. Correlated F alone
does not improve the overall comparison. Plaice has mixed metric changes:
correlated F improves F agreement but worsens SSB, N and recruitment scale
agreement. These results support keeping the options available, not replacing
the accepted stock recipes automatically.

```{r residual-summary}
residuals <- do.call(rbind, lapply(names(trials), function(name) {
  fit <- trials[[name]]$fit
  if (is.null(fit)) return(NULL)
  do.call(rbind, lapply(c("catch", "index"), function(type) {
    d <- fit$obs_pred[[type]]
    data.frame(Trial = name, Observation = type,
      Mean = mean(d$std_res, na.rm = TRUE), SD = sd(d$std_res, na.rm = TRUE),
      `Largest absolute residual` = max(abs(d$std_res), na.rm = TRUE), check.names = FALSE)
  }))
}))
knitr::kable(residuals, digits = 2)
```

Conditional residual spread also reflects fitted-state shrinkage. Persistent
survey outliers and fixed-parameter correlation advisories remain in these
applications. The dashboards retain uncertainty ribbons, new SD profiles and
the SSB-timing table; no unavailable uncertainty is manufactured.

## Validation

The full package suite passed 1,965 assertions with one interactive-only skip.
All 32 comparative-assessment test scripts and general database validation
passed. `R CMD check --no-manual` completed with zero errors, warnings or notes.
These checks include density normalization, negative/zero age correlation,
age-SD alignment, spawning-time initialization and parent recruitment,
projections, warm starts, updates, folds, reporting and dashboard rendering.
