---
title: "Catchability formula effects: recovery and stock experiments"
output:
  html_document:
    self_contained: true
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 120, fig.path = "results/q_figures/")
library(ggplot2)
study <- readRDS("results/q_recovery.rds")
stopifnot(isTRUE(study$complete))
stocks <- readRDS("results/q_stocks/results.rds")
workflow <- readRDS("results/q_workflow.rds")
d <- study$diagnostics
p <- study$parameters
p$kind <- ifelse(grepl("^sd_", p$parameter), "Process SD",
  ifelse(grepl("^phi_", p$parameter), "AR1 correlation",
    ifelse(grepl("^q_a50_", p$parameter), "Logistic midpoint", "Logistic slope")))
labels <- c(iid = "IID by survey", rw = "RW by survey", ar1 = "AR1 by survey",
  intercept = "Random intercept", numeric_by = "RW multiplied by age", logistic = "Logistic curve")
summarize <- function(data, fields, fun) {
  keys <- interaction(data[fields], drop = TRUE, lex.order = TRUE)
  do.call(rbind, lapply(split(data, keys), function(x) cbind(x[1, fields, drop = FALSE], fun(x))))
}
numerical <- summarize(d, c("scope", "spec", "link", "design"), function(x) {
  fitted <- is.finite(x$gradient)
  data.frame(attempts = nrow(x), fitted = sum(fitted), passed = sum(x$converged),
    pass_percent = round(100 * mean(x$converged), 1),
    curve_RMSE = round(mean(x$q_rmse[x$converged], na.rm = TRUE), 3),
    curve_coverage = round(100 * mean(x$q_coverage[x$converged], na.rm = TRUE), 1))
})
dense <- subset(numerical, design == "replicated")
full <- subset(dense, scope == "full")
```

## Main findings

The study completed `r sum(d$scope == "isolated")` isolated attempts and
`r sum(d$scope == "full")` full-assessment attempts. Well-informed full-model
numerical success ranged from `r min(full$pass_percent)`% to
`r max(full$pass_percent)`%. The planned 90% target was
`r if (all(full$pass_percent >= 90)) "met in every well-informed full-model case" else "not met in every well-informed full-model case"`.

The formulas are mathematically checked, but numerical convergence is not proof
that changing catchability is distinguishable from abundance or observation error.
Sparse cases and individual stock results below determine where caution is needed.
No tolerances were relaxed, no unsuccessful fits were rescued, and no settled
stock specifications were replaced.

With replicated observations and known abundance, every Gaussian case passed;
logistic success was 87% with the log link and 73% with the logit link. Full-model
Gaussian success was 83–100%, and logistic success was 73–77%. Process SD recovery
was generally reasonable with replication, but estimated AR1 correlation tended
to be below the generating value. These results support cautious use, not a
claim that every option is numerically reliable.

## What was tested

Each effect was tested with both log and logit links: IID effects by survey,
RW and AR1 trajectories by survey, conventional random intercepts, a shared RW
multiplied by age/8, and separate logistic age curves for two surveys. Process
SDs and AR1 correlation were estimated, not fixed. Logistic midpoints and slopes
were also estimated. The generating SD was 0.18 and AR1 correlation 0.65.

The isolated experiment treats abundance and survival as known. It uses 40 years,
eight ages and two surveys, with observation log-SD 0.1 estimated alongside q.
There are `r study$n_isolated` replicates per effect, link and support design.
Sparse Gaussian designs have one observation per year and survey; sparse logistic
designs see only ages 1–3, without the upper plateau.

The full assessment has 30 years, ages 1–8, two surveys at year fractions 0.25
and 0.75, fixed M=0.25, exponential initial abundance, RW recruitment, and IID F
around an age-specific mean. Catch log-SD is fixed at 0.08; survey log-SD 0.1
is estimated. There are `r study$n_full` replicates per effect/link, plus sparse
IID and AR1 cases. Starts retain the generating population values and baseline q,
but reset formula states, SDs, correlations and logistic parameters. These are
favorable-start recovery tests, not a test of arbitrary starting values.

Only the catchability formula changes in stock experiments. GOA pollock replaces
its fixed annual ADF&G term with an IID/RW/AR1 effect restricted to that survey.
Other candidates add temporal q variation or replace the age curve with logistic
selectivity. Canonical data, settled baselines and other model settings remain intact.

## Numerical performance and q recovery

A pass requires optimizer code zero, successful uncertainty calculation with a
positive-definite Hessian, and maximum absolute gradient ≤0.01. Rejections before
fitting remain in the denominator. Curve RMSE and interval coverage below use
numerically successful fits, so their denominators are smaller when fitting fails.
Coverage is first calculated within each replicate and then averaged; age/year
cells are not independent experimental replicates.

```{r numerical-figure}
ggplot(numerical, aes(labels[spec], pass_percent, fill = design)) +
  geom_col(position = "dodge") + geom_hline(yintercept = 90, linetype = 2) +
  facet_grid(scope ~ link) + coord_flip() +
  labs(x = NULL, y = "Numerical passes (%)", fill = "Support") + theme_bw()
```

Sparse full-assessment trials cover only IID and AR1. Other blank sparse bars
were not tested; zero-success log-link IID/random-intercept cases were rejected.

```{r numerical-table}
knitr::kable(numerical, row.names = FALSE)
```

```{r recovery-figure, fig.height=8}
pd <- merge(p, d[, c("scope", "spec", "link", "design", "replicate", "converged")],
  by = c("scope", "spec", "link", "design", "replicate"))
sd_data <- subset(pd, design == "replicated" & kind == "Process SD" & converged)
ggplot(sd_data, aes(labels[spec], estimate, fill = scope)) +
  geom_boxplot(outlier.shape = NA) + geom_hline(yintercept = .18, linetype = 2) +
  facet_wrap(~ link, ncol = 1) + coord_flip() +
  labs(x = NULL, y = "Estimated process SD", fill = "Experiment") + theme_bw()
```

<details>
<summary>Parameter recovery and interval coverage</summary>

```{r recovery-table}
recovery <- summarize(pd, c("scope", "spec", "link", "design", "kind"), function(x) {
  x <- x[x$converged & is.finite(x$estimate) & is.finite(x$lower) & is.finite(x$upper), ]
  if (!nrow(x)) return(data.frame(replicates = 0, bias = NA, coverage_percent = NA, MC_SE = NA))
  within <- summarize(x, "replicate", function(z) data.frame(
    bias = mean(z$estimate - z$truth), coverage = mean(z$lower <= z$truth & z$upper >= z$truth)))
  data.frame(replicates = nrow(within), bias = round(mean(within$bias), 3),
    coverage_percent = round(100 * mean(within$coverage), 1),
    MC_SE = round(100 * sd(within$coverage) / sqrt(nrow(within)), 1))
})
knitr::kable(recovery, row.names = FALSE)
```

</details>

MC_SE is the Monte Carlo standard error of coverage in percentage points.
Small differences from 95% are not interpreted without this sampling uncertainty.
Zero MC_SE when every interval covers does not establish perfect coverage.

```{r example-curves, fig.height=8}
curves <- do.call(rbind, study$examples[grepl("^full_.*_replicated_1$", names(study$examples))])
logistic_curves <- subset(curves, spec == "logistic" & year == 2000)
curves <- subset(curves, age == 4 & survey == "early" & spec != "logistic")
ggplot(curves, aes(year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#93C25C", alpha = .35) +
  geom_line(aes(y = truth), color = "#003E56") +
  geom_line(aes(y = estimate), color = "#F05C28") +
  facet_grid(spec ~ link, scales = "free_y") + labs(y = "q", x = "Year") + theme_bw()
```

Dark lines show generating q; orange lines are fitted q and ribbons are fitted
95% intervals. These are first-replicate examples, not selected best fits.

```{r logistic-curves}
ggplot(logistic_curves, aes(age)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#93C25C", alpha = .35) +
  geom_line(aes(y = truth), color = "#003E56") +
  geom_line(aes(y = estimate), color = "#F05C28") +
  facet_grid(survey ~ link) + labs(y = "q", x = "Age") + theme_bw()
```

Logistic effects change the age curve rather than adding a temporal trajectory.

## Sparse support and identifiability

```{r sparse-errors}
errors <- subset(d, !is.finite(gradient))
if (nrow(errors)) knitr::kable(summarize(errors,
  c("scope", "spec", "link", "design", "message"), function(x) data.frame(attempts = nrow(x))))
```

One observation per IID level is an exact process/observation-variance alias
under the log link and is rejected. A nonlinear logit link does not make sparse
data automatically informative: inspect variance recovery, coverage and warnings.
Logistic curves without observations near the plateau can trade off their maximum
q, midpoint and slope. Prefer supported age ranges, simpler age curves or
externally justified fixed process SDs; numerical success alone is insufficient.
Sparse numeric-by RW experiments had downward-biased process SDs despite 100%
isolated numerical success. A small multiplier can leave the shared trajectory
weakly measured. The sparse trials therefore provide a counterexample to using
convergence alone as the acceptance criterion.

## Stock applications

All five settled baselines converged. Ten of fourteen candidates converged.
EBS pollock RW/AR1 failed the gradient check; Southern Gulf cod RW had a very
large gradient and failed uncertainty estimation. Herring logistic had a small
gradient but a non-positive-definite Hessian. GOA cod, GOA pollock, Southern Gulf
AR1 and herring RW/AR1 passed numerical checks. None replaces its settled baseline.

```{r stock-table}
knitr::kable(stocks$diagnostics[, c("assessment_id", "model", "is_converged",
  "optimizer_code", "max_abs_gradient", "positive_definite_hessian", "objective")], digits = 4)
```

<details>
<summary>Candidate process and selectivity parameter intervals</summary>

```{r stock-parameters}
stock_parameters <- do.call(rbind, stocks$parameters)
stock_parameters <- merge(stock_parameters,
  stocks$diagnostics[, c("assessment_id", "model", "is_converged")],
  by = c("assessment_id", "model"))
knitr::kable(stock_parameters[, c("assessment_id", "model", "is_converged",
  "par", "coef", "est", "lwr", "upr")], digits = 3, row.names = FALSE)
```

</details>

```{r stock-comparison}
s <- subset(stocks$summaries, metric %in% c("ssb", "recruitment"))
knitr::kable(s[, c("assessment_id", "model", "metric", "comparison_status",
  "median_absolute_percent_difference", "trend_correlation")], digits = 3)
```

Percent differences describe scale agreement; trend correlation describes
trajectory agreement. Source timing, age and biomass-definition differences remain
as labelled by the canonical comparison workflow. Better agreement is not evidence
that a candidate's extra process is identifiable or scientifically preferable.

Dashboards for each stock are in `results/q_stocks/<assessment_id>/dashboard.html`.
They show Accepted, the settled baseline and candidate fits, including clearly
labelled numerical failures. Process intervals and conditional residuals should
be reviewed alongside the diagnostics rather than selecting the closest trajectory.

## Updates, projections, retrospectives and simulation

```{r workflow-table}
knitr::kable(workflow, row.names = FALSE)
```

These integration runs use estimated q-process SDs and AR1 correlation. They
check compact-state names after peeling, historical terminal years with three
existing projection years, one-year hindcast projections, and simulation from
the projected fit. A successful partial retrospective is reported with its actual
number of retained folds; failed folds are not counted as successes.

## Scope of the decision

Retain the formula options as experimental tools with the existing exact-alias
checks. Routine permission to specify a formula is separate from evidence that a
particular stock supports it. Use parsimonious q effects, inspect SD/correlation
uncertainty, and test sensitivity against the settled baseline before interpreting
time-varying q. This work does not authorize replacing any accepted stock recipe
or adding variance processes. The tables, warnings and dashboards are the review
gate for stock-specific decisions.

An observation-SD derivative bug found by the study was fixed: concatenating fixed
catch SDs before estimated survey SDs had discarded their differentiation type.
The correction preserves the existing SD formulas and observation likelihood.
The report and scripts are tracked; fits, figures and rendered output are ignored.
