---
title: "Recruitment self-tests, cross-tests, and noise"
output:
  html_document:
    self_contained: true
    toc: false
    mathjax: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 120, fig.path = "results/recruitment_cross_figures/")
library(dplyr)
library(ggplot2)
labels <- c(rw = "RW", bh_iid = "BH + IID", bh_ar1 = "BH + AR1",
  ricker_iid = "Ricker + IID", ricker_ar1 = "Ricker + AR1")
cross <- readRDS("results/recruitment_cross_recovery.rds")
noise <- readRDS("results/recruitment_noise_recovery.rds")
stopifnot(isTRUE(cross$complete), isTRUE(noise$complete))
flatten <- function(study) do.call(c, lapply(study$results, `[[`, "results"))
fits <- flatten(cross)
attempts <- do.call(rbind, lapply(fits, `[[`, "attempt"))
noise_fits <- flatten(noise)
noise_attempts <- do.call(rbind, lapply(noise_fits, `[[`, "attempt"))
stopifnot(nrow(attempts) == 25L * (cross$n_simple + cross$n_full),
  nrow(noise_attempts) == 6L * (noise$n_simple + noise$n_full),
  !anyDuplicated(with(attempts, paste(stage, truth, fitted, replicate))))
attempts$near_RW <- vapply(fits, function(x) !is.null(x$estimates$logit_phi_r) &&
  plogis(x$estimates$logit_phi_r) > .98, logical(1))
attempts$support_warning <- with(attempts, curve_warning | persistence_warning | near_RW)
correlation <- do.call(c, lapply(cross$results, function(dataset) {
  vapply(dataset$results, function(fit) {
    p <- fit$estimates
    if (is.null(p)) return(NA_real_)
    case <- fit$attempt$fitted
    R <- if (fit$attempt$stage == "isolated") dataset$generation$log_R else c(p$log_r0, p$log_r)
    if (case == "rw") e <- diff(R) else if (fit$attempt$stage == "full") {
      e <- log(fit$pairs$recruitment) - log(fit$pairs$prediction)
    } else {
      S <- dataset$generation$S[-length(R)]
      density <- p$log_sr_beta + log(S)
      mean <- p$log_sr_alpha + log(S) - if (startsWith(case, "bh"))
        pmax(0, density) + log1p(exp(-abs(density))) else exp(density)
      e <- R[-1L] - mean
      if (endsWith(case, "ar1")) e[-1L] <- e[-1L] - plogis(p$logit_phi_r) * e[-length(e)]
    }
    if (endsWith(case, "ar1")) e <- e[-1L]
    if (length(e) < 15L || any(!is.finite(e)) || sd(e) == 0) return(NA_real_)
    cor(head(e, -1L), tail(e, -1L))
  }, numeric(1))
}))
attempts$innovation_correlation <- unname(correlation)
```

## What is being tested?

Generating recruitment from BH does not guarantee that a short assessment can
recover its shape. In the original full BH–IID example, parent SSB varied only
1.7-fold. Its true median recruitment changed by about 26%; the log-scale
variation in that median was 0.052, against an innovation SD of 0.35. That is
a weak signal relative to annual recruitment variation. In that sample, the
correlation between log parent SSB and log recruitment was -0.03 in the
simulated population and -0.02 in the fitted states. The fitted curve was
almost flat, despite numerical convergence: a monotone BH curve cannot follow
a negative association. The dashboard displayed the fitted result correctly.
The fitted recruitment SD was 0.366 (95% interval 0.300–0.447), close to the
generating 0.35; it was not grossly inflated.

This study asks two separate questions:

- Can the correct model recover recruitment parameters, and what happens when
  the wrong recruitment model is fitted?
- Does lower recruitment SD help, or is a wider range of parent SSB needed?

## Equation audit

The implemented log curves are `log(alpha) + log(S) - log(1 + beta*S)`
for BH and `log(alpha) + log(S) - beta*S` for Ricker. Independent checks
of the ordinary-scale equations, parameter derivatives, derivatives with
respect to log SSB, parent-year alignment, Gaussian likelihoods and displayed
curves agreed to numerical precision. Simulation uses the parent SSB from
the same chronological population. IID SD is the residual SD; AR1 SD is the
innovation SD, with a stationary first residual. No lognormal mean correction
is applied: both curves describe median recruitment.

```{r original-contrast}
original <- readRDS("results/recruitment_recovery.rds")
original_contrast <- do.call(rbind, lapply(names(labels)[-1L], function(case) {
  x <- original$results[[paste0("full_", case, "_well informed_1")]]
  p <- setNames(x$parameters$truth, sub(":1$", "", x$parameters$parameter))
  S <- head(x$recruitment$ssb, -1L)
  beta <- exp(p[["log_sr_beta"]])
  g <- if (startsWith(case, "bh")) exp(p[["log_sr_alpha"]]) * S / (1 + beta * S) else
    exp(p[["log_sr_alpha"]]) * S * exp(-beta * S)
  data.frame(Model = labels[[case]], SSB_ratio = max(S) / min(S),
    Median_recruitment_ratio = max(g) / min(g), Log_curve_SD = sd(log(g)),
    Recruitment_SD = exp(p[["log_sd_r"]]),
    Sample_correlation = cor(log(S), log(tail(x$recruitment$truth, -1L))))
}))
knitr::kable(original_contrast, digits = 3)
```

These are the first prespecified examples, not a ranking of curve families.
The Ricker examples did **not** have greater SSB contrast or a stronger
recruitment signal than BH–IID. Their sampled SSB straddled the Ricker peak,
so a shallow dome could fit despite a weak negative association. BH must
increase, whereas Ricker can decline: their derivatives with respect to
log SSB are `1/(1 + beta*S)` and `1 - beta*S`, respectively. A visually
plausible Ricker curve alone does not demonstrate better identification.
Both full designs initialize beta as `1 / initial SSB` and pass the curve
through initial recruitment. This puts Ricker at its peak at initial SSB,
while BH remains increasing there; it is not an equal-curvature experiment.
The displayed curves also extend below the sampled SSB range. Much of the
visible Ricker rise is extrapolation. A nearly flat Ricker segment can occur
near a finite peak, while BH approaches a flat segment through its plateau;
similar-looking fits can therefore have very different parameter uncertainty.
For BH, a constant positive recruitment limit is approached as alpha and beta
both grow with their ratio held fixed. That ratio can be well estimated while
the two parameters are not. This is a boundary of the curve family, not an
equation error. An intercept with `iid(year)` is a useful simpler comparison
when the data favor nearly constant recruitment.

The first experiment crosses **RW, BH–IID, BH–AR1, Ricker–IID and Ricker–AR1**:
every simulated dataset is fitted with all five options. There are
`r cross$n_simple` isolated datasets and `r cross$n_full` full-assessment datasets
per truth, giving `r nrow(attempts)` requested fits. Matching truth/fitted entries
are self-tests; off-diagonal entries are cross-tests. RW truth has no
stock–recruit curve parameters to recover.

Isolated tests observe 80 annual recruitments directly, with parent SSB spanning
200–5,000. Curves use alpha 2, beta 0.001, SD 0.35 and AR1 correlation 0.6.
RW draws accumulate annual log increments independently of the supplied SSB.
That isolated SSB series has an ordered time trend and can act as a proxy
for time when a wrong curve is fitted. In full RW populations, SSB is itself
affected by earlier recruitment. The cross-tests therefore illustrate specific
ways to obtain misleading curves, rather than calibrating a general test.
The full assessments use the earlier 50-year, eight-age, two-survey design.
Recruitment and parent SSB come from one chronological simulated population.
Mortality, biology and observation assumptions are identical across the five
fits to each dataset.

All full candidates start from one prespecified RW fit to that dataset, even
if its numerical checks fail but it returns estimates. Its status is recorded.
An error in that starting fit prevents candidate fits and is retained. This
differs from the original recovery study's rule of stopping at a failed warm
start. No candidate receives a rescue optimization or altered tolerance.
Isolated curves start through first recruitment at first parent SSB, with
beta = 1/SSB, SD 0.5 and AR1 correlation 0.5. Numerical success requires
optimizer code zero, a positive-definite Hessian, and maximum gradient <0.01.

## Numerical success and support warnings

Across all cross-tests, 2,498/2,500 isolated fits and 698/750 full fits passed
the stated numerical checks. Correct full-model self-tests passed 28/30 RW,
28/30 BH–IID, 29/30 BH–AR1, 25/30 Ricker–IID and 28/30 Ricker–AR1.
Ricker–IID fell below the 90% numerical target. Its five failures had optimizer
code zero and positive-definite Hessians, but gradients of 0.0102–0.0179.
The threshold was not relaxed and these fits remain failures.

```{r cross-summary, include=FALSE}
summary <- attempts |>
  group_by(stage, truth, fitted) |>
  summarise(requested = n(), converged = sum(success),
    success_percent = 100 * mean(success),
    support_warning_percent = if (any(success)) 100 * mean(support_warning[success]) else NA_real_,
    correlation_warning_percent = if (any(success)) 100 * mean(parameter_correlation[success]) else NA_real_,
    residual_persistence_percent = if (any(success)) 100 * mean(abs(innovation_correlation[success]) > .5, na.rm = TRUE) else NA_real_,
    .groups = "drop")
```

```{r convergence}
ggplot(summary, aes(fitted, truth, fill = success_percent)) +
  geom_tile() + geom_text(aes(label = round(success_percent)), size = 3.5) +
  facet_wrap(~ stage) + scale_x_discrete(limits = names(labels), labels = labels) +
  scale_y_discrete(limits = rev(names(labels)), labels = labels) +
  scale_fill_gradient(limits = c(0, 100), low = "white", high = "steelblue") +
  labs(x = "Fitted recruitment model", y = "Generating model", fill = "Success (%)") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r warnings}
ggplot(subset(summary, fitted != "rw"), aes(fitted, truth, fill = support_warning_percent)) +
  geom_tile() + geom_text(aes(label = round(support_warning_percent)), size = 3.5) +
  facet_wrap(~ stage) + scale_x_discrete(limits = names(labels)[-1L], labels = labels) +
  scale_y_discrete(limits = rev(names(labels)), labels = labels) +
  scale_fill_gradient(limits = c(0, 100), low = "white", high = "darkorange") +
  labs(x = "Fitted curve", y = "Generating model", fill = "Support warnings (%)") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

Warnings are counted only among numerically successful fits. They include
parent SSB varying by less than twofold, a curve-parameter interval spanning
more than tenfold, an AR1 correlation interval wider than 0.5, or correlation
above 0.98. The isolated tests evaluate recruitment-specific advisories;
full fits use `check_tam()`, also retaining covariance and generic parameter
correlation checks. These are descriptive safeguards, not calibrated tests
for the presence of a biological stock–recruit relationship.

The detailed table also explores correlation between consecutive recruitment
innovations, flagging absolute correlation above 0.5 with at least 15 values.
The first stationary AR1 residual is excluded. This is a separate exploratory
screen evaluated from the saved attempts. It is now also a `check_tam()`
advisory for stock-recruit models: persistent unexplained recruitment can
reveal an unsuitable IID assumption even when curve intervals look precise.
The recorded support-warning rates above remain separate from this new screen.
Full-assessment innovations depend on inferred states and their shrinkage,
so this statistic is also descriptive rather than a formal test.

<details>
<summary>Counts and diagnostic rates for every truth/fitted combination</summary>

```{r counts}
knitr::kable(summary, digits = 1)
```

```{r failures}
failure_summary <- attempts |> filter(!success) |> group_by(stage, truth) |>
  summarise(failed = n(), optimizer_failure = sum(is.na(optimizer) | optimizer != 0),
    Hessian_failure = sum(!pdHess), gradient_failure = sum(!is.finite(gradient) | gradient >= .01),
    failed_warm_start = sum(!warm_start_converged, na.rm = TRUE), .groups = "drop")
knitr::kable(failure_summary)
```

Failure categories can overlap. A failed warm start is retained even when a
candidate subsequently passes; these columns do not imply every candidate
failure was caused by its starting fit. Full optimizer messages, gradients,
warnings and returned estimates remain in the saved attempts.

</details>

## Parameter recovery in compatible models

```{r coverage}
parameters <- do.call(rbind, lapply(fits, function(x) {
  if (is.null(x$parameters)) return(NULL)
  p <- x$parameters
  names(p)[names(p) == "truth"] <- "generating_value"
  cbind(x$attempt[rep(1L, nrow(p)), c("stage", "truth", "fitted", "success")], p)
}))
coverage <- parameters |>
  filter(success, is.finite(lower), is.finite(upper)) |>
  group_by(stage, truth, fitted, parameter) |>
  summarise(intervals = n(), coverage_percent = 100 * mean(lower <= generating_value & generating_value <= upper),
    median_error = median(estimate - generating_value), median_interval_width = median(upper - lower), .groups = "drop")
```

Curve recovery is evaluated only when generating and fitted curves belong to
the same family. Residual SD and correlation recovery are evaluated only for
self-tests, because their meanings differ under the wrong residual process.
Intervals are assessed on the optimized log/logit scale. There is no invented
alpha or beta truth for RW-generated recruitment.

All 500 isolated self-tests passed numerically, but nominal 95% curve-parameter
coverage ranged from 85% to 98%. Full Ricker–IID beta intervals covered truth
in only 19/25 passing fits (76%). Full BH parameter coverage was 100%, largely
because its intervals were extremely wide; that is not evidence of precise
recovery. A plausible-looking curve therefore does not establish either
identification or reliable interval calibration. Keep these options
experimental and investigate profiles and broader SSB contrast before relying
on curve-parameter intervals.

<details>
<summary>Parameter coverage and interval widths</summary>

```{r coverage-table}
knitr::kable(coverage, digits = 2)
```

</details>

## BH: noise versus parent-SSB contrast

The second experiment fits BH–IID to BH–IID data at SDs 0.15, 0.35 and 0.60
(roughly 15%, 36% and 66% CV). It contains `r nrow(noise_attempts)` requested fits.
Isolated SSB is either concentrated near 1,000 or spans 200–5,000. In full
assessments, the natural design is compared with a known oscillating fishing
pressure covariate. That covariate enters both generating and fitted F means,
creating SSB variation independently of recruitment noise. This is a declared
synthetic contrast experiment, not a modification to any stock recipe.

```{r noise-summary}
noise_contrast <- do.call(rbind, lapply(noise$results, function(x) {
  a <- x$results$bh_iid$attempt
  S <- head(x$generation$S, -1L)
  p <- x$generation$truth
  baseline <- p$log_sr_alpha + log(S) - log1p(exp(p$log_sr_beta) * S)
  data.frame(stage = a$stage, contrast = a$contrast, sigma = a$sigma,
    parent_SSB_ratio = max(S) / min(S), log_curve_signal = sd(baseline),
    signal_to_noise = sd(baseline) / a$sigma)
}))
noise_summary <- noise_attempts |>
  group_by(stage, contrast, sigma) |>
  summarise(requested = n(), converged = sum(success), success_percent = 100 * mean(success),
    support_warning_percent = if (any(success)) 100 * mean(curve_warning[success]) else NA_real_, .groups = "drop") |>
  left_join(noise_contrast |> group_by(stage, contrast, sigma) |>
    summarise(across(c(parent_SSB_ratio, log_curve_signal, signal_to_noise), median), .groups = "drop"),
    by = c("stage", "contrast", "sigma"))
knitr::kable(noise_summary, digits = c(0, 0, 2, 0, 0, 1, 1, 1, 3, 2))
```

SSB ratios and signal measures are medians over all generated datasets. The
signal is the SD of the true log median recruitment over sampled parent SSB;
signal-to-noise divides that by the generating recruitment SD. These describe
the experimental contrast, rather than a formal identifiability threshold.

```{r noise-curves}
examples <- Filter(function(x) {
  a <- x$results$bh_iid$attempt
  a$stage == "full" && a$replicate == 1L
}, noise$results)
curves <- do.call(rbind, lapply(examples, function(x) {
  fit <- x$results$bh_iid
  if (is.null(fit$curve)) return(NULL)
  S <- fit$curve$ssb
  true <- exp(x$generation$truth$log_sr_alpha) * S /
    (1 + exp(x$generation$truth$log_sr_beta) * S)
  rbind(data.frame(contrast = fit$attempt$contrast, sigma = fit$attempt$sigma,
    ssb = S, recruitment = true, curve = "Generating median"),
    data.frame(contrast = fit$attempt$contrast, sigma = fit$attempt$sigma,
      ssb = S, recruitment = fit$curve$est, curve = "Fitted median"))
}))
points <- do.call(rbind, lapply(examples, function(x) {
  data.frame(contrast = x$results$bh_iid$attempt$contrast,
    sigma = x$results$bh_iid$attempt$sigma,
    ssb = head(x$generation$S, -1L), recruitment = exp(x$generation$log_R[-1L]))
}))
ggplot(curves, aes(ssb, recruitment)) +
  geom_point(data = points, colour = "grey60", alpha = .5, size = 1) +
  geom_line(aes(colour = curve), linewidth = .7) +
  facet_grid(contrast ~ sigma, scales = "free") +
  labs(x = "Parent SSB", y = "Recruitment", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

The six examples above are the first prespecified full-assessment replicates,
including any unsuccessful fits that still returned a curve. Points are
simulated parent SSB and recruitment; lines are the generating and fitted
median relationships. These examples illustrate the sampled SSB range; the
tables retain every replicate rather than selecting successful examples.

```{r noise-parameters}
noise_parameters <- do.call(rbind, lapply(noise_fits, function(x) {
  if (is.null(x$parameters)) return(NULL)
  cbind(x$attempt[rep(1L, nrow(x$parameters)), c("stage", "contrast", "sigma", "success")], x$parameters)
}))
ggplot(subset(noise_parameters, success & grepl("log_sr_", parameter)),
  aes(factor(sigma), estimate - truth, fill = contrast)) +
  geom_hline(yintercept = 0, colour = "grey60") + geom_boxplot() +
  facet_grid(parameter ~ stage, scales = "free_y") +
  labs(x = "Recruitment innovation SD", y = "Curve-parameter error on log scale", fill = "SSB contrast") + theme_minimal()
```

Lowering SD also reduces recruitment-driven SSB variation in a closed
population. It need not improve shape recovery if it removes the source of
parent-SSB contrast. Useful curve information requires both contrast and an
adequate signal relative to noise. Local Wald intervals can be misleading near
nearly flat curve limits; wide parameter intervals and numerical failures
remain part of the evidence.

In the completed SD-0.35 experiment, wider contrast gave a median parent-SSB
range of 10.3-fold, versus 1.76-fold in the natural design. The median beta
interval width fell from 8.58 to 1.21 on the log scale. Numerical success was
28/30 versus 29/30: substantially better curve information did not require a
higher convergence rate. All 180 full BH noise fits had adjacent-innovation
correlation below the new advisory threshold.

All seeds, warnings, errors, attempted fits and available uncertainty remain
in the ignored result objects. None are removed because a wrong model fits
well or a correct model fits poorly. A converged curve, or absence of a
warning, does not establish a causal stock–recruit relationship.
