---
title: "Recruitment formulas: recovery and practical limits"
output:
  html_document:
    self_contained: true
    toc: false
    mathjax: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 120, fig.path = "results/recruitment_figures/")
study <- readRDS("results/recruitment_recovery.rds")
stopifnot(isTRUE(study$complete))
results <- study$results
attempts <- do.call(rbind, lapply(results, `[[`, "attempt"))
stopifnot(nrow(attempts) == 14L * study$n_simple + 6L * study$n_full,
  !anyDuplicated(with(attempts, paste(stage, case, design, replicate))))
library(ggplot2)
library(dplyr)
parameters <- do.call(rbind, lapply(results, function(x) {
  if (is.null(x$parameters)) return(NULL)
  cbind(x$attempt[rep(1L, nrow(x$parameters)), c("stage", "case", "design", "replicate", "success")], x$parameters)
}))
parameters$covered <- with(parameters, lower <= truth & truth <= upper)
summary <- attempts |>
  group_by(stage, case, design) |>
  summarise(attempts = n(), passed = sum(success),
    success_percent = round(100 * mean(success), 1),
    median_log_curve_error = round(median(curve_rmse, na.rm = TRUE), 3), .groups = "drop")
```

## Main findings

Recruitment formulas let us separate an environmental or spawning-biomass
relationship from unexplained good and bad years. Numerical convergence,
recovery of the relationship, and useful uncertainty are separate questions.
The comparisons below do not rank stock assessments or establish a biological
relationship for a particular stock.

The study records `r nrow(attempts)` unique attempts, including
`r sum(attempts$stage == "full")` full assessments.
`r sum(!attempts$success)` did not pass all numerical criteria. Every warning,
error and seed is retained. Failed fits are not rescued or removed.

All 1,400 isolated attempts passed. Full assessments passed in 30/30 BH–IID,
29/30 BH–AR1, 28/30 Ricker–IID, 27/30 Ricker–AR1, 24/30 covariate–IID and
28/30 covariate–RW attempts. The stock–recruit cases reached the numerical
target; covariate–IID did not. Four of its six failures came from the
prespecified RW warm start; the remaining two had gradients just above 0.01.
No failed attempt was rerun with adjusted starts or tolerances.

Retain the new options experimentally, with the default RW unchanged.
Numerical success is encouraging, but it does not establish density dependence.
The local curve intervals and curve-parameter coverage are not uniformly
reliable, particularly with narrow SSB support or extrapolation.

## Model and study design

Beverton–Holt rises toward a plateau; Ricker can decline at high parent SSB.
Both describe **median recruitment**, with positive productivity and
density-dependence parameters. IID deviations are independent good and bad
years; AR1 deviations persist between years. Covariate RW models instead
explain changes in log recruitment. The default RW is unchanged.

- **Isolated recovery:** 100 attempts per case with 80 annual recruitment
  values observed directly. Four curve/process combinations and two covariate
  cases are tested. Each curve also has narrow-SSB and SSB-correlated-covariate
  stress cases. Starts are a declared 0.1 offset from the generating parameters;
  this tests the density and information in the design, not assessment starts.
- **Full assessments:** 30 attempts per case, 50 years, eight ages, two surveys,
  fixed M, age-specific IID F, and no additional cohort-survival deviations.
  Recruitment is simulated chronologically from the same population that
  produces catch and survey observations. The fixed initial recruitment is
  1,000 fish; recruitment innovation SD is 0.35 and AR1 correlation is 0.6.
  A prespecified data-based RW fit supplies common starting values. Failure of
  that warm start is counted as failure of the attempt, without a rescue fit.

The criterion is optimizer code zero, positive-definite fixed-effect Hessian,
and maximum absolute gradient below 0.01. Coverage is conditional on numerical
success and a finite interval; its denominator is shown. Parameter intervals
are assessed on their optimized log/logit/coefficient scale. Curve error is
root mean squared error on the log scale. Full stock–recruit curves use a common
SSB grid extending beyond the observed range; recruitment-state recovery is
retained separately in the raw results.

The full assessments are rich in observations, but their closed population
dynamics can produce a narrow parent-SSB range or concentrate observations
near a curve's plateau. They are not guaranteed to be rich in information
about density dependence. This distinction matters when reading their
numerical success and curve-parameter coverage.

One initial execution completed the isolated runs but stopped before the full
fits because the worker-launch call was invalid. That checkpoint was retained;
the corrected execution repeated the isolated runs with identical seeds. No
model, data, starting rule or tolerance was changed in response to fit failure.
Documentation and guards for unsupported extreme lags/saturated short series
were finalized during the study; neither changes any study model's equations,
inputs or fitting rule. The checkpoint retains the source fingerprint at launch.

A subsequent review found a study-only observation assignment error: survey
draws in prepared row order were assigned to unsorted input rows. The 132
completed full attempts and three interrupted attempts were retained in
`results/recruitment_invalid_observation_order`, with the original study code,
and excluded from scientific inference. All 180 full attempts were restarted
with the correct survey/year/age assignment. A regression test now checks
round-trip observation alignment. The 1,400 isolated results were retained:
their function body and density/input rules were verified unchanged. Earlier
full-assessment recovery claims from the invalid run should be disregarded.

## Numerical success and curve recovery

```{r numerical}
knitr::kable(summary, digits = 3)
```

```{r numerical-figure}
ggplot(summary, aes(case, success_percent, fill = design)) +
  geom_col(position = "dodge") + facet_wrap(~ stage) +
  geom_hline(yintercept = 90, linetype = 2) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = NULL, y = "Numerical success (%)", fill = NULL) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The dashed line is the prespecified 90% target in well-informed cases. Stress
cases can converge while providing little information about curve shape.

<details>
<summary>Recorded numerical failures</summary>

```{r failures}
knitr::kable(attempts[!attempts$success,
  c("stage", "case", "replicate", "optimizer", "gradient", "pdHess", "message")], digits = 4)
```

</details>

## Representative fitted relationships

```{r curve-data, include=FALSE}
curves <- do.call(rbind, lapply(results, function(x) {
  if (is.null(x$curve) || x$attempt$stage != "full" || x$attempt$replicate != 1L ||
      startsWith(x$attempt$case, "cov")) return(NULL)
  cbind(case = x$attempt$case, x$curve)
}))
```

```{r fitted-curves}
ggplot(curves, aes(ssb, truth)) +
  geom_line(colour = "grey30", linewidth = .8) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = .15) +
  geom_line(aes(y = estimate), colour = "steelblue", linewidth = .8) +
  facet_wrap(~ case, scales = "free") +
  labs(x = "Parent SSB", y = "Median recruitment") + theme_minimal()
```

Grey is the generating curve; blue is the first recorded fitted curve with its
95% confidence interval. These intervals describe the median relationship,
not the spread of future recruitment or uncertainty in parent SSB.

```{r curve-coverage}
curve_intervals <- do.call(rbind, lapply(results, function(x) {
  if (!isTRUE(x$attempt$success) || is.null(x$curve$lower)) return(NULL)
  d <- x$curve
  usable <- is.finite(d$lower) & is.finite(d$upper)
  in_range <- if (is.null(x$recruitment)) rep(TRUE, nrow(d)) else
    d$ssb >= min(x$recruitment$ssb) & d$ssb <= max(x$recruitment$ssb)
  covered <- d$lower <= d$truth & d$truth <= d$upper
  cbind(x$attempt[, c("stage", "case", "design", "replicate")],
    available = mean(usable),
    coverage_in_range = mean(covered[usable & in_range]),
    coverage_beyond_range = mean(covered[usable & !in_range]))
}))
curve_coverage <- curve_intervals |>
  filter(stage == "full") |>
  group_by(case) |>
  summarise(fits = n(), finite_intervals_percent = round(100 * mean(available), 1),
    coverage_in_range_percent = round(100 * mean(coverage_in_range), 1),
    coverage_beyond_range_percent = round(100 * mean(coverage_beyond_range), 1),
    .groups = "drop")
knitr::kable(curve_coverage)
```

Coverage averages over the common grid within each successful fit, then over
fits. The range is the generating population's historical parent SSB. Curve
intervals use a local Gaussian approximation. A nearly flat fitted BH curve
can have narrow prediction intervals despite extremely uncertain productivity
and density dependence; those intervals need not cover alternative shapes,
especially beyond observed SSB. Do not treat them as reliable extrapolation.

## Parameter recovery and coverage

<details>
<summary>Parameter coverage and interval widths</summary>

```{r coverage}
coverage <- parameters |>
  filter(success, is.finite(lower), is.finite(upper)) |>
  group_by(stage, case, design, parameter) |>
  summarise(intervals = n(), coverage_percent = round(100 * mean(covered), 1),
    median_error = round(median(estimate - truth), 3),
    median_interval_width = round(median(upper - lower), 3), .groups = "drop")
knitr::kable(coverage)
```

</details>

```{r parameter-errors}
ggplot(subset(parameters, stage == "full" & success),
  aes(case, estimate - truth)) +
  geom_hline(yintercept = 0, colour = "grey60") + geom_boxplot() +
  facet_wrap(~ parameter, scales = "free_y") +
  labs(x = NULL, y = "Error on optimized parameter scale") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

High coverage with very wide intervals is not precise recovery. Narrow parent
SSB or correlated environmental effects can hide density dependence. In full
assessments, estimated parent SSB and mortality add uncertainty that is absent
from isolated recovery.

## Workflow and interpretation

```{r workflow}
workflow <- readRDS("results/recruitment_workflow.rds")
knitr::kable(workflow[, setdiff(names(workflow), "warnings")])
```

The workflow exercises projections, year-aligned starts, retrospective and
hindcast folds, generative simulations, tidy uncertainty, and dashboards.
Warnings and fold convergence are retained independently of structural checks.
All three projections and dashboards passed. One BH retrospective fold and
one covariate-RW hindcast fold failed numerical convergence and were excluded
by the existing fold helpers; their warnings remain recorded. This is not a
claim that every fold converges.
Stock–recruit density excludes early fixed boundary states; unavailable
pre-model parent SSB is never invented. Projection covariates stay at their
terminal values. IID residuals forecast toward zero; AR1 residuals decay
toward zero at their fitted correlation.

Start with a simpler recruitment model and examine the range of parent SSB,
curve intervals, residuals, and sensitivity to environmental covariates before
interpreting density dependence. The options remain experimental. This study
does not validate overlapping recruitment processes, steepness, reference
points, spawning-time SSB, or arbitrary covariate extrapolation.
