---
title: "Observation SD effects: age patterns and recovery"
output:
  html_document:
    self_contained: true
    toc: false
    mathjax: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 120, fig.path = "results/sd_figures/")
results <- readRDS("results/sd_effects_recovery.rds")
stopifnot(isTRUE(results$complete))
attempts <- results$attempts
stopifnot(nrow(attempts) == 14L * results$n_simple + 6L * results$n_full,
  !anyDuplicated(with(attempts, paste(stage, type, design, replicate))))
parameters <- results$parameters
library(ggplot2)
success <- aggregate(success ~ stage + type + design, attempts, mean)
counts <- aggregate(replicate ~ stage + type + design, attempts, length)
names(counts)[names(counts) == "replicate"] <- "attempts"
success <- merge(success, counts)
parameters <- merge(parameters, attempts[, c("stage", "type", "design", "replicate", "success")])
parameters$covered <- with(parameters, lower <= truth & truth <= upper)
curves <- do.call(rbind, lapply(results$curves, function(d) {
  if (!"group" %in% names(d)) d$group <- "catch"
  d
}))
```

## Main findings

Random age effects can describe changing observation spread without estimating
an unrelated SD at every age. IID effects partially pool age SDs; RW and AR1
allow neighbouring ages to be similar. They **do not require** noisier young
and old fish. A fixed quadratic age curve remains a useful simpler choice when
that pattern is plausible.

The study separates numerical success from recovery. A fit can pass numerical
checks yet recover a poor SD curve or an imprecise correlation. Start with
replicated age effects, inspect uncertainty, and compare simpler SD formulas.

- **IID age effects are a useful starting point.** All 30 catch and 30 survey
  fits passed numerical checks. Survey process SD was reasonably recovered;
  catch process SD tended to be underestimated (median about 0.26 versus 0.35).
- **Smoothing can work, but depends on the observation type.** Survey RW passed
  28/30 fits and survey AR1 passed 29/30; catch AR1 passed 30/30. AR1 correlations
  remained imprecise and tended to be underestimated. Many intervals were so
  wide that high coverage did not imply useful precision.
- **Catch-SD RW needs particular caution.** Only 13/30 full fits passed;
  17 stopped with non-finite gradients. All well-replicated isolated RW fits
  passed, so this is a full-assessment numerical problem, not evidence of an
  incorrect RW density. An advisory now flags this specification.
- **Do not overlook fixed curves.** All random-effect curves captured the
  noisier tails, but the quadratic curve had the smallest error when the truth
  was quadratic. Sparse random SDs often passed numerical checks while
  recovering poor curves or underestimating process SD.

Treat temporal changes in SD as exploratory: this study tests processes over
**age**, not stochastic volatility over time or formulas for latent N/F/M SDs.

There are `r nrow(attempts)` recorded attempts, including
`r sum(attempts$stage == "full")` full assessments.
`r sum(!attempts$success)` did not pass the numerical criteria. Errors and
warnings are retained; there are no rescue fits or relaxed tolerances.

## What changes in the model?

The expected log observation is unchanged. Its SD is

`log(SD) = log(supplied SD) + fixed effects + random effects`

A positive effect increases residual spread; negative effects reduce it.
The process SD describes variation in **log SD**. For example, an effect of
0.3 multiplies observation SD by about 1.35. This process SD is distinct from
the observation SD and the F/M/N process SDs. Formula effects appear under
Catch SD or Survey SD on the Random parameters page, on the log-SD scale.
The baseline SD is the value when the effect is zero; it is not the arithmetic
average of the varying SDs.

Useful starting formulas are `~ iid(age)` for catch and
`~ survey + rw(age, by = survey)` for survey SD. Categorical `by` gives separate
age curves with a shared process SD/correlation; numeric `by` multiplies one
shared effect. Supplied SDs remain offsets. RW starts at zero and its first
age's baseline comes from ordinary formula terms. AR1 has a stationary start.

## Study design

The isolated experiment observes Normal log residuals with known zero mean,
15 ages, baseline SD 0.25, effect-process SD 0.35 and AR1 correlation 0.65.
There are 30, three, or one observations per age, and
`r results$n_simple` replicates per specification. Baseline SD, process SD and
AR1 correlation are estimated. Sparse random SDs produce a scale mixture;
they are not an exact alias of Gaussian random effects on the mean.
RTMB uses a Laplace approximation for these non-Gaussian random-SD likelihoods.

The noisy-tail experiment uses the same 30 observations per age and the true
curve `exp(log(0.18) + 0.025 * (age - 8)^2)`. It compares a common SD, quadratic
fixed curve, IID, RW and AR1 with the **same observations**. The quadratic model
matches the generating shape by design; this is not a general ranking.

Full assessments have 30 years, 15 ages, catch at age and two surveys at different
times. N deviations are off, F has IID residuals with age-specific mean, M is
supplied at 0.25, and q has separate survey baselines. The study varies **one**
SD surface at a time: catch age effects or survey-specific age effects. Other
SDs are estimated fixed levels. Recruitment, F, q and both observation-error
levels are estimated; starts use the generating fixed baselines, without
knowledge of the simulated latent states. There are `r results$n_full`
replicates per specification. Simultaneously varying q, N and SD over the
same ages/years is not validated here.

Numerical success requires optimizer code zero, a positive-definite uncertainty
Hessian, finite predictions/SDs for full assessments, and maximum absolute
gradient below 0.01. The well-informed target is 90% success. Success alone
does not establish structural identifiability or reliable intervals.

## Numerical checks

```{r numerical}
ggplot(success, aes(type, 100 * success, fill = design)) +
  geom_col(position = "dodge") + facet_wrap(~ stage, scales = "free_x", nrow = 1) +
  geom_hline(yintercept = 90, linetype = 2, colour = "grey40") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = NULL, y = "Numerically successful fits (%)", fill = NULL) + theme_bw()
```

```{r success-table}
summary <- success
summary$success <- round(100 * summary$success, 1)
names(summary)[names(summary) == "success"] <- "Success (%)"
knitr::kable(summary, row.names = FALSE)
```

## Recovery of SD and correlation

The distributions below include finite estimates from failed fits; hollow red
points identify those fits. Zero Hessian uncertainty, missing intervals and
non-finite estimates are not interpreted as successful recovery.

```{r parameters, fig.height=9}
ggplot(parameters[is.finite(parameters$estimate), ], aes(type, estimate)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(data = parameters[!parameters$success & is.finite(parameters$estimate), ],
    colour = "firebrick", shape = 1, position = position_jitter(width = .1)) +
  geom_hline(aes(yintercept = truth), data = unique(parameters[, c("stage", "design", "parameter", "truth")]),
    colour = "steelblue", linetype = 2) +
  facet_wrap(~ stage + design + parameter, scales = "free", ncol = 3) +
  labs(x = NULL, y = "Estimated parameter; blue line = truth") + theme_bw()
```

```{r coverage}
valid <- parameters[parameters$success & is.finite(parameters$lower) & is.finite(parameters$upper), ]
coverage <- aggregate(covered ~ stage + type + design + parameter, valid, mean)
counts <- aggregate(replicate ~ stage + type + design + parameter, valid, length)
names(counts)[names(counts) == "replicate"] <- "Intervals"
coverage <- merge(coverage, counts)
coverage$covered <- round(100 * coverage$covered, 1)
names(coverage)[names(coverage) == "covered"] <- "Coverage (%)"
knitr::kable(coverage, row.names = FALSE)
```

Coverage is conditional on numerical success and available intervals. Short
age series provide few independent process states even when each age has many
years of observations. AR1 correlation can therefore remain imprecise.

## Recovering the noisier tails

The grey line is the simulated SD; colours show fitted curves in the first
replicate, using the same observations for each formula.

```{r tails}
tail_curves <- subset(curves, stage == "tails")
ggplot(tail_curves, aes(age)) + geom_line(aes(y = truth), colour = "grey35", linewidth = 1) +
  geom_line(aes(y = estimate, colour = type)) +
  labs(x = "Age", y = "Observation SD", colour = "Formula") + theme_bw()
```

```{r curve-errors}
rmse <- attempts[is.finite(attempts$log_sd_rmse), ]
ggplot(rmse, aes(type, log_sd_rmse, fill = design)) +
  geom_boxplot(outlier.shape = NA) + facet_wrap(~ stage, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = "Error in the fitted log-SD curve (RMSE)", fill = NULL) + theme_bw()
```

## Full-model examples

These are the first replicate of every full-model specification, selected
before fitting. Returned curves remain shown when numerical checks fail.
The first catch-RW fit stopped before producing estimates, so it has no curve;
it remains in the failed-fit table.

```{r full-curves}
full <- subset(curves, stage == "full")
full <- merge(full, attempts[, c("stage", "type", "design", "replicate", "success")])
full$panel <- paste(full$design, full$type, ifelse(full$success, "passed", "did not pass"))
ggplot(full, aes(age, colour = group)) +
  geom_line(aes(y = truth), linetype = 2) + geom_line(aes(y = estimate)) +
  facet_wrap(~ panel, scales = "free_y", ncol = 3) +
  labs(x = "Age", y = "Observation SD; dashed = simulated truth", colour = NULL) + theme_bw()
```

<details><summary>Warnings and failed fits</summary>

```{r failures}
fail <- attempts[!attempts$success, c("stage", "type", "design", "replicate", "seed",
  "optimizer", "gradient", "pdHess", "message")]
knitr::kable(fail, row.names = FALSE)
warnings <- unique(attempts[nzchar(attempts$warnings), c("stage", "type", "design", "warnings")])
knitr::kable(warnings, row.names = FALSE)
```

</details>

## Scope and reproducibility

```{r workflow}
if (file.exists("results/sd_workflow.rds")) {
  knitr::kable(readRDS("results/sd_workflow.rds"), row.names = FALSE)
}
```

Workflow checks use the first catch IID and survey RW examples. They retain
failed-fold warnings and actual fold counts, check stable state names, simulate
from returned states, and render dashboards with log-SD effects and intervals.

Retain one ordered SD process per observation surface. Saturated fixed/random
SD designs and duplicate IID variance components are rejected; sparse support
and wide SD/correlation intervals are advisory findings, not automatic failure.
No stock recipes, latent-process mathematics, or convergence criteria were
changed for this study. These experiments are deliberately informative and
do not establish performance for arbitrary sparse assessments.

Run `simulate_sd_effects.R 100 30 3` from the repository root and render this
report. Seeds, every attempted fit and session information are saved in the
ignored `results/sd_effects_recovery.rds`. Only code, report text and
deterministic tests are tracked. Generated figures, dashboards and fits stay
local.
The optional third argument controls independent workers; checkpoints resume
missing attempts with the same explicit replicate seeds. Worker scheduling
does not change the generating observations or model settings.

```{r study-notes, results='asis'}
if (length(results$study_notes)) cat(results$study_notes)
```
