Diagnostics
================================================================================

Row {.tabset}
--------------------------------------------------------------------------------

### Numerical checks

```{r, results = "asis"}
for (nm in names(checks)) {
  cat("#### ", nm, "\n\n", sep = "")
  print(knitr::kable(checks[[nm]]$numerical))
  cat("\n\n")
}
```

### Structural checks

Passing these checks does not establish that the data distinguish abundance,
mortality and catchability, or that the biological assumptions are suitable.

```{r, results = "asis"}
for (nm in names(checks)) {
  cat("#### ", nm, "\n\n", sep = "")
  print(knitr::kable(checks[[nm]]$structural))
  cat("\n\n")
}
```

### Advisory findings

These findings flag patterns to investigate. They do not change numerical
convergence. Conditional residuals also reflect fitted-state shrinkage.

```{r, results = "asis"}
for (nm in names(checks)) {
  cat("#### ", nm, "\n\n", sep = "")
  print(knitr::kable(checks[[nm]]$advisories))
  cat("\n\n")
}
```

### Residual summaries

Only positive historical observations are included. Lag-1 correlation uses
adjacent years, with at least five pairs.

```{r}
knitr::kable(stack_list(lapply(checks, `[[`, "residuals"), label = "model"), digits = 3)
```

### Data support

```{r}
knitr::kable(stack_list(lapply(checks, `[[`, "data"), label = "model"))
```
